Uncertainty in social learning evolution

### 2022-05-21 meanNetPayoffs.tex
updated mean payoffs, step figs to use minipages and standalone

\begin{figure}
  \caption{Dependence of average net payoff normalized by lifespan,
    $\meanpi/L$, at the
    end of last time step of the simulation on environmental
    variability $u$ for several values of low payoff $\pilow$, number of
    behaviors $B$, and lifespan $L$. Dashed lines mark expected 
    payoff from individual learning $\langle \pi_I \rangle$ calculated
    for each parameter setting (not sensitive to $u$). 
    Note that when the maximum $L$ is 
    8, the y-axis runs from 0 to 8, and when the maximum $L$ is 20, the y-axis
    runs from 0 to 20. Average over 1000 trials.} 
  \label{fig:meanNetPayoffs}
  \centering
  \begin{subfigure}[t]{0.08\textwidth}
    \begin{tikzpicture}
      \draw[->,thick] (-.1,0)--(1,0) node[above]
        {$B$};
      \draw[->,thick] (0,.1)--(0,-1) node[below]
        {$\pilow$};
    \end{tikzpicture}

  \end{subfigure}%
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    {\large $B = 2$} 
    \begin{rotate}{90}
      {\parbox{4cm}{
          \centering
          \vspace{-3.5em}{\large $ \pilow = 0.1$} \\
          {\begin{rotate}{-90}{\large $\frac{\meanpi}{L}$}\end{rotate}}
      }}
    \end{rotate}%
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=2.pdf}
    }
    \centering
    \begin{rotate}{90}
      {\parbox{4cm}{
          \centering
          \vspace{-3.5em} {\large$ \pilow = 0.45$} \\
          {\begin{rotate}{-90}{\large $\frac{\meanpi}{L}$}\hspace{1em}\end{rotate}}
      }}
    \end{rotate}%
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=2.pdf}
    }
    \begin{rotate}{90}
      {\parbox{4cm}{
          \centering
          \vspace{-3.5em} {\large $ \pilow = 0.8$} \\
          {\begin{rotate}{-90}{\large $\frac{\meanpi}{L}$}\end{rotate}}
      }}
    \end{rotate}%
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.8_nbehaviors=2.pdf}
    }\\[-.5em]
    {\large $\quad u$}
  \end{subfigure}%
  \begin{subfigure}[t]{0.3\textwidth}
    {\large $B = 4$}
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=4.pdf}
    } \\
    \centering
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=4.pdf}
    } \\
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.8_nbehaviors=4.pdf}
    }\\[-.5em]
    {\large{$\quad u$}}
  \end{subfigure}
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    {\large $B = 10$}
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=10.pdf}
    } \\
    \centering
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=10.pdf}
    } \\
    \includegraphics[width=1.1\textwidth]{
      {Figures/mean_prev_net_payoff_over_u_lowpayoff=0.8_nbehaviors=10.pdf}
    } \\[-.5em]
    {\large $\quad u$}
  \end{subfigure}

\end{figure}

>>>>

\documentclass[varwidth=true,crop=false]{standalone}
\usepackage[chatter]{rotating}
\usepackage{amssymb,amsmath}
\usepackage{pgfplots}

\usepackage{geometry}
\geometry{
paperwidth=8in,
paperheight=6.25in,
margin=0.25in
}

\newcommand{\pisub}[1]{\pi_{\mathrm{#1}}}
\newcommand{\pilow}{\pisub{low}}
\newcommand{\pihigh}{\pisub{high}}
\newcommand{\piI}{\langle \pisub{I} \rangle}
\newcommand{\piS}{\langle \pisub{S} \rangle}
\newcommand{\ledger}{\bar\pi_{ib}}

\newcommand{\meanvar}[1]{\langle #1 \rangle}
\newcommand{\meansl}{\meanvar{s}}
\newcommand{\meanpi}{\meanvar{\pi}}
\newcommand{\meansoc}{\meanvar{\pi_\mathrm{S}}}
\newcommand{\meanasoc}{\meanvar{\pi_\mathrm{A}}}
\newcommand{\meanT}{\meanvar{T}}

\newcommand{\bandit}{\text{Bandit}_b(0, 1)}

\begin{document}

    \begin{minipage}{3.75in}
      \centering
      {\hspace{5.25em}\huge $B = 4$}
    \end{minipage}%
    \begin{minipage}{3.75in}
      \centering
      {\hspace{2.0em}\huge $B = 10$}
    \end{minipage}~\\

    \begin{minipage}{3.75in}
    \begin{rotate}{90}
      {\parbox{2.5in}{
          \centering
          \vspace{-2.5em} {\huge$ \pilow = 0.1$} \\
          {\begin{rotate}{-90}{\huge $\frac{\meanpi}{L}$}\hspace{3em}\end{rotate}}
      }}
    \end{rotate}%
    \hspace{2em}
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=4.pdf}
    \end{minipage}\noindent\begin{minipage}{3.75in}%
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.1_nbehaviors=10.pdf}
    \end{minipage}~\\[0.5em]

    \begin{minipage}{3.75in}
    \begin{rotate}{90}
      {\parbox{2.5in}{
          \centering
          \vspace{-2.5em} {\huge$ \pilow = 0.45$} \\
          {\begin{rotate}{-90}{\huge $\frac{\meanpi}{L}$}\hspace{3em}\end{rotate}}
      }}
    \end{rotate}%
    \hspace{2em}
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=4.pdf}
        \\[-2.75em]
        \begin{center}
          {\hspace{3.25em} \huge $\quad u$}
      \end{center}
        \end{minipage}\noindent\begin{minipage}{3.75in}%
      \includegraphics[width=\textwidth]{Figures/mean_prev_net_payoff_over_u_lowpayoff=0.45_nbehaviors=10.pdf}
      \\[-2.75em]
      \begin{center}
        {\huge $\quad u$}
      \end{center}
    \end{minipage}
\end{document}

### 2022-05-21 ms.tex
updated mean payoffs, step figs to use minipages and standalone

\documentclass[letterpaper,11.5pt]{scrartcl}
\usepackage{totcount}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage[chatter]{rotating}
\usepackage{amssymb,amsmath}
\usepackage[title]{appendix}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
% \usepackage[margin=0.50in]{geometry}
\usepackage[margin=1.00in]{geometry}

\usepackage{booktabs}
\usepackage{pslatex}
\usepackage{apacite}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{pgfplots}
\usepackage{wrapfig}
\usepackage[english]{babel}
\usepackage{lmodern}
\usepackage{setspace}
\doublespace
% \usepackage{url}
\usepackage{bigfoot}
\usepackage[export]{adjustbox}
\setlength\intextsep{0pt}

@@

To gain a more systematic understanding of uncertainty and social learning evoluiton,
we identified four common classes of ecological
uncertainty (``principal components'' of uncertainty) used as factors
in social learning evolution models: (1) environmental variability; (2) payoff
ambiguity (difference between best behavior choice and others); (3) 
selection set size (number of ecologically possible behaviors); and 
(4) the lifespan of agents. These were chosen because modeling and empirical 
study designs across taxa tend use one or more of these uncertainty dimensions.
Environmental variability is sometimes modeled as catastrophic and 
happens only once in an individual's lifetime~\cite{Rogers1988}, or may be
modeled as occurring with a fixed or variable
probability~\cite{Feldman1996,McElreath2005}---in these examples environments
can only be in one of two states. Payoff ambiguity is sometimes in the form
of differences in expected payoffs arising from different
behaviors or strategies~\cite{Enquist2007,Rendell2010}
or by varying the standard deviation of payoffs~\cite{McElreath2005}. 
Selection set size is often limited to one size throughout social learning
evolution studies: empirical studies of \emph{bombus terresteris} (bumble
bee)~\cite{Baracchi2018} and \emph{parus major} (great tit)~\cite{Aplin2017}
populations each provided two behaviors for
the bees and birds to perform, each yielding greater or smaller
payoffs depending on experimental treatment; studies in humans have used two
or three possible behaviors~\cite{McElreath2005,Toyokawa2019}; m
odeling studies have used either two 
behaviors~\cite{Feldman1996,Rendell2010} or four behaviors~\cite{Enquist2007} 
in understanding the role of uncertainty. The age or lifespan of individuals was
found to affect on social learning in both \emph{parus
major}~\cite{Aplin2017} and \emph{poecilia reticulata} 
(guppies)~\cite{Leris2016} \mt{Need better connection between lifespan and 
age, or better references to support use of lifespan}.
Shorter lifespans mean individuals
die more uncertain of which behaviors are optimal, and pass that uncertainty on
to their offspring. To answer our research questions about the combined effect
of various forms of uncertainty on social learning evolution, we developed an
agent-based model that incorporates all four of
these uncertainty variables. We systematically vary these parameters
to understand and predict their effects on social learning evolution.

\paragraph{Social learning evolution} (CRISTINA, PAUL)
To add to confusion about how to integrate diverse models of social learning
evolution, different models select different social learning components in
seemingly \emph{ad hoc} ways 
\begin{itemize}
  \item 
    Vertical, oblique, horizontal transmission
  \item
    Conformity, payoff-biased, frequency-dependent, etc. A note about how
    ``conformity'' is often not distinguished from other forms of social learning
  \item
    Review human studies
  \item
    Review non-human studies~\cite{Leris2016,Aplin2017,Avargues-Weber2018,Baracchi2018}
  \item
    One sentence on how dual-inheritance theory enables us to gloss over whether
    social learning is genetically or culturally evolved.
  \item
    More things from which we selected the operation of our model
\end{itemize}

\paragraph{Meaning of uncertainty in ecology and cognitive and social sciences} (JAMIE, PAUL)
We need to support our claim that these four classes of uncertainty really 
count as forms of uncertainty and show how it is consistent across ecology and
cognitive and social science.

\paragraph{Evolutionary ABMs for theory development} (PAUL, MATT)

\paragraph{Research overview}

@@

In between generations agents first reproduce (asexual haploid reproduction); 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off. $N$ reproducers are selected with replacement, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
Child agents inherit their parents social learning trait $s_i$ without mutation.
Child agents with $s_i = 1$ then learn from one teacher from their parent's
generation, including possibly their parent. Child social learners select
their teacher first by selecting $N_T$ prospective teachers from the 
population completely at random, then selecting the one with highest
net payoffs $\pi_i$ for as their teacher, with ties broken randomly 
($N_T = 5$ in the main text;
\mt{Other values will be tested in the supplement}). While it is somewhat
arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally
guaranteed. Social learners adopt their chosen teacher's ledger of 
expected payoffs $\ledger$, and set
all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$. 
Asocial learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and
$\ledger = 0$. The newly initialized population then begins behavior selection
and payoff accumulation for another $L$ time steps and the evolutionary process
continues until $\sum_i s_i = 1$ or $\sum_i s_i = 0$ \mt{I believe
all trials will eventually end like this, so in the next set of model runs I will
not allow early termination before one of these conditions are met}.

To analyze the effect of the four principal uncertainty factors, we systematically
varied their values and observed three key outcome variables aggregated across 1000
trial simulations, and across all agents. Specifically we varied $u \in \{0.0,
0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L
\in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  We observed
three outcome measures: (1) $\meansl$, the average value of $s_i$ over all agents
and trials; (2) $\meanpi / L$, the average net payoffs within agent lifetimes over
trials, normalized by their lifespan; and (3) $\meanT / L$, or the mean number
of time steps to fixation normalized by lifespan\footnote{Equivalent to the mean
number of generations to fixation} 
(fixation means $\sum_i s_i = 0 \text{ or } 1$). \mt{Make sure to qualitatively
introduce these and their evolutionary importance in the intro.}
We inspect the outcomes through visualization environmental
variation is on the x-axis since it is the best understood: we expect that as $u$
increases, social learning prevalence $\meansl \to 0$. We expected and observed
that the other three uncertainty factors shift at which $u$ $\meansl$ starts to
decrease, and how rapidly $\meansl$ decreases as $u$ increases. Note the
hypothesis-testing concept of significance is meaningless here because we could
make any small outcome difference significant by running more simulation trials.
The important outcomes for this study are the patterns of our outcome 
variables over the systematic variation of the uncertainty factors.

@@

\paragraph{Disc. 1: Foundation for studying ssortative social learning}~\cite{Katsnelson2014}.

\paragraph{Disc. 2: Applications to artificial
intelligence}~\cite{Sandholm1996,Jaques2019,Ndousse2021,Gronauer2022}

>>>>

\documentclass[letterpaper,11.5pt]{scrartcl}
\usepackage{totcount}
% \documentclass[11pt]{report}
% \documentclass{report}
% \documentclass{book}
\usepackage[bookmarks, hidelinks]{hyperref}
\usepackage{amssymb,amsmath}
\usepackage[title]{appendix}
% \usepackage{fullpage}
\usepackage{tabulary}
\usepackage{tabularx}
\usepackage{float}
% \usepackage[margin=0.50in]{geometry}
\usepackage[margin=1.00in]{geometry}

\usepackage{booktabs}
\usepackage{pslatex}
\usepackage{apacite}
\usepackage{caption}
% \usepackage{subcaption}
\usepackage{wrapfig}
\usepackage[english]{babel}
\usepackage{lmodern}
\usepackage{setspace}
\doublespace
% \usepackage{url}
\usepackage{bigfoot}
\usepackage[export]{adjustbox}
\setlength\intextsep{0pt}

@@

\mt{This paragraph is a first try to introduce the four uncertainty dimensions
with examples to illustrate the problem we are answering.
I outlined the next paragraph to get into more details of social learning:
how cultural evolution works; conformity, success-biased, etc. Cristina, please
edit this.} 
To gain a more systematic understanding of uncertainty and social learning
evoluiton, we identified four common classes of ecological
uncertainty (``principal components'' of uncertainty) used as factors
in social learning evolution models: (1) environmental variability; (2) payoff
ambiguity (difference between best behavior choice and others); (3) 
selection set size (number of ecologically possible behaviors); and 
(4) the lifespan of agents. These were chosen because modeling and empirical 
study designs across taxa tend use one or more of these uncertainty dimensions.
Environmental variability is sometimes modeled as catastrophic and 
happens only once in an individual's lifetime~\cite{Rogers1988}, or may be
modeled as occurring with a fixed or variable
probability~\cite{Feldman1996,McElreath2005}---in these examples environments
can only be in one of two states. Payoff ambiguity is sometimes in the form
of differences in expected payoffs arising from different
behaviors or strategies~\cite{Enquist2007,Rendell2010}
or by varying the standard deviation of payoffs~\cite{McElreath2005}. 
Selection set size is often limited to one size throughout social learning
evolution studies: empirical studies of \emph{bombus terresteris} (bumble
bee)~\cite{Baracchi2018} and \emph{parus major} (great tit)~\cite{Aplin2017}
populations each provided two behaviors for
the bees and birds to perform, each yielding greater or smaller
payoffs depending on experimental treatment; studies in humans have used two
or three possible behaviors~\cite{McElreath2005,Toyokawa2019}; 
modeling studies have used two 
behaviors~\cite{Feldman1996,Rendell2010} and four behaviors~\cite{Enquist2007} 
in understanding the role of uncertainty. The age or lifespan of individuals was
found to affect on social learning in both \emph{parus
major}~\cite{Aplin2017} and \emph{poecilia reticulata} 
(guppies)~\cite{Leris2016} \mt{Need better connection between lifespan and 
age, or better references to support use of lifespan}.
Shorter lifespans mean individuals
die more uncertain of which behaviors are optimal, and pass that uncertainty on
to their offspring. To answer our research questions about the combined effect
of various forms of uncertainty on social learning evolution, we developed an
agent-based model that incorporates all four of
these uncertainty variables. We systematically vary these parameters
to understand and predict their effects on social learning evolution.

\paragraph{Social learning evolution} (CRISTINA)
To add to confusion about how to integrate diverse models of social learning
evolution, different models select different social learning components in
seemingly \emph{ad hoc} ways 
\begin{itemize}
  \item 
    Vertical, oblique, horizontal transmission
  \item
    Conformity, payoff-biased, frequency-dependent, etc. A note about how
    ``conformity'' is often not distinguished from other forms of social learning
  \item
    Review human studies
  \item
    Review non-human studies~\cite{Leris2016,Aplin2017,Avargues-Weber2018,Baracchi2018}
  \item
    One sentence on how dual-inheritance theory enables us to gloss over whether
    social learning is genetically or culturally evolved.
  \item
    More things from which we selected the operation of our model
\end{itemize}

\paragraph{Meaning of uncertainty in ecology and cognitive and social sciences} (JAMIE)
We need to support our claim that these four classes of uncertainty really 
count as forms of uncertainty and show how it is consistent across ecology and
cognitive and social science.

\paragraph{ABMs for social learning evolution theory development} (PAUL)

@@

In between generations agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off. $N$ reproducers are selected with replacement, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
Child agents inherit their parents social learning trait $s_i$ without mutation.
Child agents with $s_i = 1$ then learn from one teacher from their parent's
generation, including possibly their parent. Child social learners select
their teacher first by selecting $N_T$ prospective teachers from the 
population completely at random, then selecting the one with highest
net payoffs $\pi_i$ for as their teacher, with ties broken randomly 
($N_T = 5$ in the main text;
\mt{Other values will be tested in the supplement}). While it is somewhat
arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally
guaranteed. Social learners adopt their chosen teacher's ledger of 
expected payoffs $\ledger$, and set
all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$. 
Asocial learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and
$\ledger = 0$. The newly initialized population then begins behavior selection
and payoff accumulation for another $L$ time steps and the evolutionary process
continues until $\sum_i s_i = 1$ or $\sum_i s_i = 0$ \mt{I believe
all trials will eventually end like this, so in the next set of model runs I will
not allow early termination before one of these conditions are met}.

To analyze the effect of the four principal uncertainty factors, we systematically
varied their values and observed three key outcome variables aggregated across 1000
trial simulations, and across all agents. We varied $u \in \{0.0,
0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L
\in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  We observed
three outcome measures: (1) $\meansl$, the average value of $s_i$ over all agents
and trials; (2) $\meanpi / L$, the average net payoffs within agent lifetimes over
trials, normalized by their lifespan; and (3) $\meanT / L$, or the mean number
of time steps to fixation normalized by lifespan\footnote{Equivalent to the mean
number of generations to fixation} 
(fixation means $\sum_i s_i = 0 \text{ or } 1$). \mt{Make sure to qualitatively
introduce these and their evolutionary importance in the intro.}
We inspect the outcomes through visualization environmental
variation is on the x-axis since it is the best understood: we expect that as $u$
increases, social learning prevalence $\meansl \to 0$. We expected and observed
that the other three uncertainty factors shift at which $u$ $\meansl$ starts to
decrease, and how rapidly $\meansl$ decreases as $u$ increases. Note the
hypothesis-testing concept of significance is meaningless here because we could
make any small outcome difference significant by running more simulation trials.
The important outcomes for this study are the patterns of our outcome 
variables over the systematic variation of the uncertainty factors.

@@

\section{Discussion}

By identifying four common uncertainty dimensions we were able to develop a model
that predicts how they might interact to foster or suppress social learning.
The model reproduced classic predictions that social learning can yield higher
payoffs than individual learning, but only up to some upper uncertainty bound.
By disambiguating and organizing various operationalizations of \emph{uncertainty},
we were able to add more detail to our theoretical understanding of which forms
of uncertainty interact in which ways to make social learning more or less
beneficial compared to asocial learning. When expected social learning payoffs
approached asocial learning payoffs the system exhibited far stronger drift; when
social or asocial learning were more obviously better the systems more consistently
evolved to the more profitable state. Note that in many cases we made modeling choices
that excluded some important mechanisms of social learning; for example we 
assumed success-biased learning, although conformity is theoretically important
as well~\cite{Muthukrishna2016a,Smaldino2018b}. Nonetheless, we believe our model
advances theoretical parsimony; and the software implementation is designed to
facilitate testing alternate modeling choices.

\paragraph{Disc. 2: Foundation for studying social learning evolution with group
structure (PAUL, JAMIE, CRISTINA)}~\cite{Katsnelson2014}.

Our modeling approach may have applications to developing social artificial 
intelligence. Much effort has gone in to developing the cognitive machinery
to realize multi-agent reinforcement learning
systems~\cite{Sandholm1996,Ndousse2021,Gronauer2022}, with some work explicitly
incorporating cultural evolutionary approaches to social learning~\cite{Jaques2019}.
However, it seems there are a lack of methods for adapting
to environmental uncertainty, or determining when social learning is actually
advantageous under different uncertainty conditions. Our model dynamics could
be used as an evolutionary algorithm to update AI agents' social learning
strategy without knowledge of the underlying uncertainty structure.
Our model could also predict the expected benefit of social learning,
which may be inherently costly in artificial multi-agent learning systems, 
and how long the translated evolutionary algorithm may take to converge.
